Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2020-11-10 manuscript_Huang.tex
conclusion is added.

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{logisticEntropy.pdf}
\caption{\small{Behavior of different statistical complexity measures for the Logistic map in dependence on the control parameter $r$. (a) bifurcation diagram, (b) the spectrum of the maximal Lyapunov exponent, (c) pattern frequency based (permutation) entropy $\mathcal{S}_O$ and (d) complexity $\mathcal{C}_O$; (e) pattern transition frequency based entropy $\mathcal{S}_W$ by the globally normalized weighting matrix $\mathbf{W}$ and (f) the corresponding complexity $\mathcal{C}_W$. (g, h) are entropy $\mathcal{S}_E$ and complexity $\mathcal{C}_E$ which are similar with (e, f) but based on node-wise out-link normalized weighting matrix $\mathbf{W}$. Several major periodic windows have been highlighted by grey background. Vertical dotted lines correspond to the cases summarized in Tab. \ref{tableLog}. } \label{fig:bifurcation}}
\end{figure*}

@@

\section{Transitions in continuous system}
The previous results have been further verified in the chaotic R\"ossler system \cite{Roessler1976}, namely, $(\dot{x}, \dot{y}, \dot{z}) = (-y-z, x+0.2y, b+z(x-5.7))$, where $b$ is the control parameter. There are some challenging problems for constructing OPTNs for time series from continuous systems, for instance, proper choice of sampling frequency and embedding parameters, which must depend on the particular time scales of the system. {\color{red} numerical details will be provided soon. }

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{rosslerEntropy.pdf}
\caption{\small{Similar with the caption of Fig. \ref{fig:bifurcation}. Behavior of different SCMs for the R\"ossler system varying the control parameter $b$. (a) bifurcation diagram, (b) the maximal Lyapunov exponent $\lambda$, (c) pattern frequency based permutation entropy $\mathcal{S}_O$ and (d) complexity $\mathcal{C}_O$; (e) pattern transition frequency based entropy $\mathcal{S}_W$ by the globally normalized weighting matrix $\mathbf{W}$ and (f) the corresponding complexity $\mathcal{C}_W$. (g, h) are entropy $\mathcal{S}_E$ and complexity $\mathcal{C}_E$ which are similar with (e, f) but based on node-wise out-link normalized weighting matrix $\mathbf{W}$. Several major periodic windows have been highlighted by grey background. Vertical dotted lines correspond to periodic windows. } \label{fig:bifurRossler}}
\end{figure*}

@@

\section{Discussions} \label{sec:con}
In summary, we have proposed to include ordinal pattern transitions into the traditional ordinal pattern frequency analysis, generalizing statistical complexity measures for nonlinear time series analysis. The pattern transition properties of the corresponding time series networks provide novel insights for permutation entropy which relies on the static pattern frequency only. In addition, we suggest two slightly different ways to consider the effects of pattern transitions which are based on (i) a globally normalized weight transition matrix $\mathbf{W}$ and (ii) node-wise normalized transition matrix. Note that the node-wise based SCMs consider both individual pattern frequencies and and the transition frequencies among them.

complex network approaches for nonlinear time series analysis \cite{ZouPR2018}.

Some open problems regarding to the ordinal pattern analysis need to be further explored. Length dependence, further numerical performance to show distinctions between different dynamics (various models of maps, long range correlated fractional Brownian motion, fractional Gaussian noise etc)

For periodic windows of the Logistic map, one may use longer data lengths, but does not improve the estimate very much.

Generalization from univariate to bivariate time series analysis will be further discussed. We noticed that this remains to be a largely untouched area for statistical complexity analysis. It would be an interesting task to show the dependence of complexity on interaction.

>>>>

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{logisticEntropy.pdf}
\caption{\small{Behavior of different statistical complexity measures for the Logistic map in dependence on the control parameter $r$. (a) bifurcation diagram, (b) the spectrum of the maximal Lyapunov exponent, (c) pattern frequency based (permutation) entropy $\mathcal{S}_O$ and (d) complexity $\mathcal{C}_O$; (e) pattern transition frequency based entropy $\mathcal{S}_W$ by the globally normalized weighting matrix $\mathbf{W}$ and (f) the corresponding complexity $\mathcal{C}_W$. (g, h) are entropy $\mathcal{S}_E$ and complexity $\mathcal{C}_E$ which are similar with (e, f) but based on node-wise out-link normalized weighting matrix $\mathbf{W}$. Several major periodic windows have been highlighted by grey background. Vertical dotted lines correspond to the cases as summarized in Tab. \ref{tableLog}. } \label{fig:bifurcation}}
\end{figure*}

@@

\section{Transitions in continuous system}
The previous results have been further verified in the chaotic R\"ossler system \cite{Roessler1976} while varying the control parameter, namely, $(\dot{x}, \dot{y}, \dot{z}) = (-y-z, x+0.2y, b+z(x-5.7))$, where $b$ is the control parameter. There are some challenging problems for constructing OPTNs for time series from continuous systems, for instance, proper choice of sampling frequency and embedding parameters, which must depend on the particular time scales of the system. To minimize the discussions on proper choices of further algorithmic parameters, we introduce Poincar\'e sections at $y=0$, $\dot{y}<0$ (for each value of $b$) and construct OPTNs from $N = 10000$ intersection points. The results have been shown in Fig. \ref{fig:bifurRossler}. 
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{rosslerEntropy.pdf}
\caption{\small{Similar with the caption of Fig. \ref{fig:bifurcation}. Behavior of different SCMs for the R\"ossler system varying the control parameter $b$. (a) bifurcation diagram ($x$-component on the Poinar\'e sections), (b) the maximal Lyapunov exponent $\lambda$, (c) pattern frequency based permutation entropy $\mathcal{S}_O$ and (d) complexity $\mathcal{C}_O$; (e) pattern transition frequency based entropy $\mathcal{S}_W$ by the globally normalized weighting matrix $\mathbf{W}$ and (f) the corresponding complexity $\mathcal{C}_W$. (g, h) are entropy $\mathcal{S}_E$ and complexity $\mathcal{C}_E$ which are similar with (e, f) but based on node-wise out-link normalized weighting matrix $\mathbf{W}$. Several major periodic windows have been highlighted by grey background. } \label{fig:bifurRossler}}
\end{figure*}
Rather similar results as Fig. \ref{fig:bifurcation} have been obtained in the chaotic R\"ossler system, identifying different dynamic transitions, where some periodic windows have been highlighted.

@@

\section{Conclusions} \label{sec:con}
In summary, we have proposed to include ordinal pattern transitions into the traditional ordinal pattern frequency analysis, generalizing statistical complexity measures for nonlinear time series analysis. The pattern transition properties of the corresponding time series networks provide novel insights for permutation entropy which relies on the static pattern frequency only. Therefore, this is another case showing the usefulness of ordinal pattern transition network for time series analysis \cite{ZouPR2018}. In addition, we suggest two slightly different ways to consider the effects of pattern transitions which are based on (i) a globally normalized weight transition matrix $\mathbf{W}$ and (ii) node-wise normalized transition matrix. Note that the node-wise based SCMs consider both individual pattern frequencies and and the transition frequencies among them. We show that these two new statistical complexity measures characterize the complexity properties of deterministic chaos successfully, far away from the lower right bottom of the complexity-entropy plane (Fig. \ref{fig:fBmNoise}). This helps to obtain deeper understandings of complexity analysis for time series of both deterministic chaos and stochastic systems, for instance, further numerical performance are required to show distinctions between different dynamics including various models of chaotic maps, long range correlated fractional Brownian motion and fractional Gaussian noise etc \cite{BorgesAMC2019,RossoPRE2007}. We have demonstrated the usefulness of SCMs by further results of time series for experimental fluid flows and ECG recordings, showing distinctive characteristics for different dynamics.

Some open problems of traditional nonlinear time series analysis should be further considered, for example, length dependence for reliable estimations of SCMs. For iterative processes of maps, one may use longer data lengths, which however does not improve the estimate very much due to the accumulation effects of computer iterations (Fig. \ref{fig:transient}). For continuous systems, together with embedding parameters of $D$ and $\tau$, the sampling frequency is another important one that needs to be further explored. Generalization from univariate to bivariate time series analysis will be further discussed. We noticed that this remains to be a largely untouched area for statistical complexity analysis. It will be an interesting task to show the dependence of complexity on interaction.
